### 2020-05-22 serial.tex
tweaks

\begin{flushleft}{
	\Large
	\textbf\newline{
		Unraveling the paradox between generation and serial intervals
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
Kaiyuan Sun\textsuperscript{2}
Benjamin M.\ Bolker\textsuperscript{3,4,5}
David Champredon\textsuperscript{6}
David J.\,D.\ Earn\textsuperscript{4,5}
Michael Li\textsuperscript{3}
Joshua S.\ Weitz\textsuperscript{7, 8}
Bryan T.\ Grenfell\textsuperscript{1,2,9}
Jonathan Dushoff\textsuperscript{3,4,5}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
\\
\textbf{3} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{4} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{6} Department of Pathology and Laboratory Medicine, University of Western Ontario, London, Ontario, Canada
\\
\textbf{7} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{8} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{9} Woodrow Wilson School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

Generation and serial intervals are under-appreciated yet important quantities for characterizing the outset of an outbreak.
Even though their definitions (and differences) were clearly laid out over a decade ago, there is still a need for clear conceptual and theoretical understanding of their roles in shaping the epidemiological dynamics.
Part of the confusion can be attributed to an apparent paradox.
By definition, both generation and serial intervals describe how infection spreads between individuals --- thus, one would expect them both to link between the epidemic growth rate $r$ and the reproduction number \RR.
However, current theory does not explain how two different distributions can estimate the same \RR from $r$ --- some studies have further demonstrated that generation and serial intervals give different estimates of \RR.
We provide an answer to the paradox by showing that \emph{forward-looking} serial intervals, which correctly estimates \RR from $r$, are not the same as ``intrinsic'' serial intervals, but instead change with $r$. 
We provide a heuristic way of addressing potential biases that can arise from not accounting for changes in these references and apply our framework to serial intervals of COVID-19.
This study demonstrates the importance of early epidemiological investigation through contact tracing for characterizing initial serial intervals and provides rationale for reassessing serial intervals, and \RR\ estimates, for COVID-19.

@@

The generation- and serial-interval distributions are key quantities in outbreak analyses.
Recent theoretical studies suggest that two distributions give different estimates of \RR;
however, both intervals, by definition, can equally describe the disease transmission at the individual level.
Here, we show that the serial-interval distribution, defined from the correct reference time and cohort, gives the same estimate of \RR as the generation-interval distribution.
We then apply our framework to serial interval data from the COVID-19 epidemic.
While our study supports the use of serial-interval distributions in estimating \RR, it also reveals several gaps in the current understanding and applications of the serial-interval distributions.

@@

The reproduction number \RR is one of the most important characteristics of an emerging epidemic such as coronavirus disease 2019 (COVID-19) \citep{majumder2020early}.
The reproduction number is defined as the average number of secondary cases caused by a primary case;
the reproduction number in a fully susceptible population --- referred to as the basic reproduction number \Ro --- allows us to predict the extent to which a disease will spread in the population and the amount of intervention necessary to eliminate disease \citep{anderson1991infectious}.
Near the beginning of an outbreak, \RR is often estimated from the observed exponential growth rate using generation- and serial-interval distributions (e.g., \cite{du2020serial, jung2020real, li2020early, zhao2020preliminary}).

The generation interval is defined as the time between when an individual (infector) is infected and when that individual infects another person (infectee);
the generation-interval distribution determines the relationship between the exponential growth rate $r$ and the reproduction number \RR \citep{anderson1991infectious, ferguson2005strategies, wallinga2007generation}.
Similarly, the serial interval is defined as the time between when an infector and an infectee become \emph{symptomatic} \citep{svensson2007note}.
While serial intervals are similar to generation intervals, previous studies have noted that, in many contexts, serial intervals are expected to have larger variances than generation intervals but have the same mean \citep{svensson2007note,klinkenberg2011correlation,te2013estimating,champredon2018equivalence};
some studies further suggested that using serial intervals can give different estimates of \RR \citep{britton2019estimation, ganyani2020estimating}.

@@

One source of confusion comes down to an apparent paradox.
When the epidemic is growing exponentially, the spread of infection can be characterized as a ``renewal process'' based on previous incidence of infection, the associated generation-interval distribution, and the average infectiousness of an infected individual.
This renewal formulation allows us to link the exponential growth rate of an epidemic $r$ with its reproduction number \RR using the generation-interval distribution \citep{wallinga2007generation}.
By definition, the serial-interval distribution describes the renewal process between symptomatic cases based on their symptom onset dates;
therefore, both generation- and serial-interval distributions should give identical estimates of  \RR based the observed epidemic growth rate $r$.
In contexts where the distributions are expected to be different, current theory has no explanation for how they might provide identical estimates of \RR from $r$ (cf. \cite{britton2019estimation}).

Here, we resolve this paradox, by showing that the relevant interval for the renewal framework is what is called the ``forward'' interval, and that the forward serial (but not generation) interval is affected by the rate of growth $r$ during the exponential-growth phase.
We develop a new framework for understanding serial intervals and show that the initial forward serial-interval distribution correctly estimates \RR from $r$.
Conversely, using inaccurately defined serial intervals or failing to account for changes in the observed serial-interval distributions over the course of an epidemic can significantly bias estimates of \RR.
We apply our framework to serial intervals of COVID-19 and lay out several principles to consider in using information about serial intervals and other epidemiological time delays in the analysis of epidemics.

@@

We first begin by describing a general framework for characterizing a distribution of time delays between two epidemiological events;
these events can be defined either within an infected individual (e.g., infection and symptom onset of an individual, the incubation period) or between infected individuals (e.g., symptom onsets of an infector and an infectee, the serial interval).
We can further divide these events into \emph{primary} and \emph{secondary} events.
When we measure an epidemiological time delay within an infected individual (e.g., the incubation period), the primary event is the event that always or usually occurs before the secondary event ---
most epidemiological events that can be observed within an individual have clear direction (e.g., infection and onset of symptoms) but some may not (e.g., onset of infectiousness and onset of symptoms).
When we measure an epidemiological time delay between infected individuals (e.g., the serial interval), 
the primary and secondary events are defined in terms of the direction of transmission:
the primary event refers to the event that occurs in the infector; again some of these delays are always positive (the infector is always infected before the infectee) and some are not (it is possible for the infectee to report symptoms before the infector).

@@

Both forward and backward perspectives
must yield identical \emph{measurement} (e.g., the length of the incubation
period of a given individual is the same whether measured forward from
the time of infection or backward from the time of symptom onset).
Consequently, if we restrict attention to individuals with a given
incubation period, say $\tau$, and we count how many became infected at time $\psymp$
or how many first showed symptoms at time $\ssymp=\psymp+\tau$ then we must obtain
the same number. In general, no matter how delays are distributed, if
$\mathcal P$ and $\mathcal S$ represent the sizes of primary and
secondary cohorts then we must have
\begin{equation}
\psize(\psymp) f_\psymp(\tau) = \ssize(\ssymp) b_\ssymp(\tau) \,.
\label{eq:match}
\end{equation}
Substituting $\psymp=\ssymp-\tau$, it follows that
\begin{equation}
b_\ssymp(\tau)= \frac{\psize(\ssymp-\tau) f_{\ssymp-\tau} (\tau)}{\ssize(\ssymp)}
\label{eq:backward}
\end{equation}
If we are thinking of realized incubation periods, then the left hand side of
this equation is the probability density that an individual
who became symptomatic at time $\ssymp$ had an incubation period of
length $\tau$. From the right hand side, we see that this probability
density depends on how many individuals became infected at time
$\ssymp-\tau$ [$\psize(\ssymp-\tau)$], how many individuals became
symptomatic at time $\ssymp$ [$\ssize(\ssymp)$], and the probability density
that someone who was infected at time $\ssymp-\tau$ became symptomatic at
time $\ssymp$ [$f_{\ssymp-\tau}(\tau)$]. In general, Eq.~\eqref{eq:backward}
shows that the the backward delay distribution at reference time $\ssymp$
[$b_\ssymp(\tau)$] depends not only on the secondary cohort size at time
$\ssymp$ [$\ssize(\ssymp)$], but also on the primary cohort size at the
earlier time $\ssymp-\tau$ [$\psize(\ssymp-\tau)$] and the forward delay
distribution at that earlier time [$f_{\ssymp-\tau}(\tau)$].

There are multiple mechanisms that drive the changes in forward and backward delay distributions over time.
Typically, within-individual forward delay distributions are not affected by epidemic dynamics.
Some, like incubation periods, are expected to be equivalent to their intrinsic distributions and remain invariant over the course of an epidemic.
Others, like the time from symptom onset to testing, may change over the course of an epidemic due to changes in public health policies.
Between-individual forward delay distributions, such as generation- or serial-interval distributions, depend on epidemic dynamics;
for example, forward generation intervals contract during the outbreak because infected individuals are less likely to infect others later due to susceptible depletion \citep{champredon2015intrinsic}.

@@

The serial interval is defined as the time between when an infector becomes symptomatic and when an infectee becomes symptomatic \citep{svensson2007note}.
Previous studies have often expressed serial intervals $\tau$ in the form (\fref{diagram}A):
\begin{equation}
\tau = (\gtime + x_1) - x_0
\end{equation}
where $x_0$ and $x_1$ represent incubation periods of an infector and an infectee, respectively, and $\gtime$ represents the generation interval.
These studies assumed that $x_0$ and $x_1$ follow the same distributions and concluded that the serial and generation intervals have the same mean \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation};
however, this conclusion is based on the implicit assumption that incubation times, $x_0$ and $x_1$, as well as generation interval, $\gtime$, are intrinsic to individuals (and not dependent on disease dynamics) --- something that is true of forward but not backward incubation times.
We refer to this definition as the intrinsic serial interval.

@@

The total number of serial intervals can now be obtained by integrating over all possible infection times for the infector and the infectee:
\begin{equation}
\total (\psymp,\ssymp) = \int_{-\infty}^{\psymp} \int_{\pinf}^{\ssymp} \Rc (\pinf) i(\pinf) h_{\pinf}(\psymp-\pinf, \sinf - \pinf) \idist_{\sinf}(\ssymp - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
Then, the forward serial-interval distribution $f_\psymp(\tau)$ is given by the density of intervals of length $\tau$ starting at \psymp, relative to the total number of serial intervals starting at \psymp: 
\begin{equation}
f_\psymp(\tau) = 
\frac{\total(\psymp, \psymp+\tau)}{\int \total(\psymp, \psymp+\tau') d\tau'}.
\end{equation}
Likewise, the backward serial-interval distribution $b_\ssymp(\tau)$ is given by the density of intervals of length $\tau$ ending at \ssymp, relative to the total number of serial intervals ending at \ssymp: 
\begin{equation}
b_\ssymp(\tau) = 
\frac{\total(\ssymp-\tau, \ssymp)}{\int \total(\ssymp-\tau', \ssymp) d\tau'}.
\end{equation}

@@

We illustrate the theory by applying it to a specific example of an epidemic model. 
We model disease spread with a renewal-equation model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be written as:
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\\
i(t) &= \Ro S(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\end{aligned}
\label{eq:renewal}
\end{equation}
where \Ro is the basic reproduction number, and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a population where the number of susceptibles remains constant).
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \citep{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) \propto \gdist(\tau) S(\psymp + \tau),
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) \propto h(x, \tau) S(\psymp + \tau).
\end{equation}
We further assume that the forward incubation period distributions does not vary across cohorts over the course of an epidemic, as they represent the life history of a disease, and use $k$ without subscript instead. 
Then, we have:
\begin{equation}
\begin{aligned}
\idist(x) &= \int_0^\infty h(x, \tau) \dtau\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx
\end{aligned}
\end{equation}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) \dtau.
\end{equation}

We use this framework to illustrate how the realized epidemiological time distributions vary over the course of an epidemic and depend on the perspective.
We run deterministic as well as stochastic simulations based on the renewal equations (see Appendix) using previously estimated intrinsic incubation period and intrinsic generation-interval distributions for COVID-19 (Table 1) and compare both the forward and backward delay distributions;
in particular, we focus on the realized incubation period, generation-interval, and serial-interval distributions.

@@

During the initial phase of an epidemic, the proportion susceptible remains approximately constant ($S(t) \approx S(0)$) and incidence of infection grows exponentially: $i(t)=i_0\exp(rt)$.
During this period, the intrinsic generation-interval distribution provides the correct link between the exponential growth rate $r$ and the initial reproduction number $\RR=\Ro S(0)$ based on the Euler-Lotka equation \citep{wallinga2007generation};
in this case, we focus on the estimates of the basic reproduction number \Ro (the value of \RR in a fully susceptible population, $S(t) \approx 1$):
\begin{equation}
\frac{1}{\Ro} = \int_0^\infty \exp(-r\tau) g(\tau) \dtau.
\end{equation}
Analogous to the intrinsic generation-interval distribution, 
forward serial-interval distributions, by definition, describe the renewal process of infection based on symptom onset dates.
Therefore, we expect the forward serial-interval distribution during the exponential growth phase --- which we refer to as the \emph{initial} forward serial-interval distribution $f_0$ --- to estimate the same value of \Ro for a given $r$ as the intrinsic generation-interval distribution (note, however, that the forward serial interval is not necessarily positive):
\begin{equation}
\frac{1}{\Ro} = \int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau,
\end{equation}
where the initial forward serial-interval distribution is given by:
\begin{equation}
f_{0}(\tau) \propto \int_{-\infty}^{0} \int_{\pinf}^{\tau} \exp(r \pinf) h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
In the Appendix, we provide a mathematical proof of this relationship.
We further compare this with the estimate of \Ro based on the intrinsic serial-interval distribution:
\begin{equation}
\frac{1}{\Rintrinsic} = \int_{-\infty}^\infty \exp(-r\tau) q(\tau) \mathrm{d} \tau,
\end{equation}
where the intrinsic serial-interval distribution does not depend on disease dynamics:
\begin{equation}
q(\tau) \propto \int_{-\infty}^{0} \int_{\pinf}^{\tau} h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}

@@

\begin{table}[!th]
\begin{center}
\begin{tabular}{|l|l|r|}
\hline
Parameter & Values & Source\\
\hline
Mean forward incubation period & 5.5 days & \cite{lauer2020incubation} \\
SD forward incubation period & 2.4 days & \cite{lauer2020incubation} \\
Mean intrinsic generation interval & 5 days & \cite{ferretti2020quantifying} \\
SD intrinsic generation interval & 2 days & \cite{ferretti2020quantifying} \\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Parameter values used for simulations.}
The intrinsic generation-interval distribution is parameterized using a log-normal distribution with log mean $\mu_G=1.54$ and log standard deviation $\sigma_G=0.37$.
The forward incubation period distribution is parameterized using a log-normal distribution with log mean $\mu_I=1.62$ and log standard deviation $\sigma_I=0.42$.
The joint probability distribution is modeled using a multivariate log-normal distribution with correlations (on the log scale) $\rho=-0.5, 0, 0.5$.
}
\end{table}

>>>>

\begin{flushleft}{
	\Large
	\textbf\newline{
		Cohort-based approach to understanding the roles of generation and serial intervals in shaping epidemiological dynamics
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
Kaiyuan Sun\textsuperscript{2}
Benjamin M.\ Bolker\textsuperscript{3,4,5}
David Champredon\textsuperscript{6}
David J.\,D.\ Earn\textsuperscript{4,5}
Michael Li\textsuperscript{3}
Joshua S.\ Weitz\textsuperscript{7, 8}
Bryan T.\ Grenfell\textsuperscript{1,2,9}
Jonathan Dushoff\textsuperscript{3,4,5}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
\\
\textbf{3} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{4} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{6} Department of Pathology and Laboratory Medicine, University of Western Ontario, London, Ontario, Canada
\\
\textbf{7} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{8} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{9} Woodrow Wilson School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

Generation intervals and serial intervals are critical quantities for characterizing outbreak dynamics.
Generation intervals characterize the time between infection and transmission; whereas serial intervals characterize the time between the onset of symptoms in a chain of transmission.
They are often used interchangeably, leading to misunderstanding of how to use either generation or serial intervals to establish a link between the epidemic growth rate $r$ and the reproduction number \RR.
On one hand, generation intervals provide a mechanistic link between $r$ and \RR but are harder to measure via contact tracing.
On the other hand, serial intervals are easier to measure from contact tracing but recent studies suggest that two intervals give different estimates of \RR from $r$.
We present a general framework for characterizing \emph{any} epidemiological delays based on cohorts (i.e., a group of individuals that share the same event time, such as symptom onset) and show that \emph{forward-looking} serial intervals, which correctly estimates \RR from $r$, are not the same as ``intrinsic'' serial intervals, but instead change with $r$.
We provide a heuristic method for addressing potential biases that can arise from not accounting for changes in serial intervals across cohorts and apply the method to estimating \RR for the COVID-19 outbreak in China using serial interval data --- our analysis shows that using incorrectly defined serial intervals can severely bias the estimate.
This study demonstrates the importance of early epidemiological investigation through contact tracing and provides rationale for reassessing generation intervals, serial intervals, and \RR\ estimates, for COVID-19.

@@

The generation- and serial-interval distributions are key, but different, quantities in outbreak analyses.
Recent theoretical studies suggest that two distributions give different estimates of the reproduction number \RR from the exponential growth rate $r$;
however, both intervals, by definition, describe disease transmission at the individual level.
Here, we show that the serial-interval distribution, defined from the correct reference time and cohort, gives the same estimate of \RR as the generation-interval distribution.
We then apply our framework to serial interval data from the COVID-19 outbreak in China.
While our study supports the use of serial-interval distributions in estimating \RR, it also reveals necessary changes to the current understanding and applications of serial-interval distribution.

@@

The reproduction number \RR is one of the most important characteristics of an emerging epidemic, including characterization of coronavirus disease 2019 (COVID-19) \citep{majumder2020early}.
The reproduction number is defined as the average number of secondary cases caused by a primary case;
the reproduction number in a fully susceptible population --- referred to as the basic reproduction number \Ro --- allows us to predict the extent to which a disease will spread in the population and the amount of intervention necessary to eliminate disease \citep{anderson1991infectious}.
Near the beginning of an outbreak, \RR is often estimated from the observed exponential growth rate using generation- and serial-interval distributions (e.g., \cite{du2020serial, jung2020real, li2020early, zhao2020preliminary}).

The generation interval is defined as the time between when an individual (infector) is infected and when that individual infects another person (infectee);
the generation-interval distribution determines the relationship between the exponential growth rate $r$ and the reproduction number \RR \citep{anderson1991infectious, ferguson2005strategies, wallinga2007generation}.
Similarly, the serial interval is defined as the time between when an infector and an infectee become \emph{symptomatic} \citep{svensson2007note}.
While serial intervals are similar to generation intervals, previous studies have noted that, in many contexts, serial intervals are expected to have larger variances than generation intervals but have the same mean \citep{svensson2007note,klinkenberg2011correlation,te2013estimating,champredon2018equivalence}.

@@

One source of confusion comes down to an apparent discrepancy between biological intuition and mathematical theory.
When the epidemic is growing exponentially, the spread of infection can be characterized as a ``renewal process'' based on previous incidence of infection, the associated generation-interval distribution, and the average infectiousness of an infected individual.
This renewal formulation allows us to link the exponential growth rate of an epidemic $r$ with its reproduction number \RR using the generation-interval distribution \citep{wallinga2007generation}.
By definition, the serial-interval distribution describes the renewal process between symptomatic cases based on their symptom onset dates;
therefore, we expect both generation- and serial-interval distributions to give identical estimates of  \RR based the observed epidemic growth rate $r$.
In contexts where the distributions are expected to be different, current theory has no explanation for how they might provide identical estimates of \RR from $r$. 
Some studies have further suggested that using serial intervals can give different estimates of \RR \citep{britton2019estimation, ganyani2020estimating}.

Here, we resolve this discrepancy, by showing that the relevant interval for the renewal framework is what is called the ``forward'' interval, and that the forward serial (but not generation) interval is affected by the rate of growth $r$ during the exponential-growth phase.
We develop a new framework for understanding serial intervals, as well as any other epidemiological delays, and show that the initial forward serial-interval distribution correctly estimates \RR from $r$.
Conversely, using inaccurately defined serial intervals or failing to account for changes in the observed serial-interval distributions over the course of an epidemic can significantly bias estimates of \RR.
We apply our framework to serial intervals of COVID-19 and lay out several principles to consider in using information about serial intervals and other epidemiological time delays to correctly infer the reproduction number during the early stages of an outbreak.

@@

The time delays between two epidemiological events can be defined either within an infected individual (e.g., infection and symptom onset of an individual, the incubation period) or between infected individuals (e.g., symptom onsets of an infector and an infectee, the serial interval).
We can further divide these events into \emph{primary} and \emph{secondary} events.
When we measure an epidemiological time delay within an infected individual (e.g., the incubation period), the primary event is the event that always or usually occurs before the secondary event ---
most epidemiological events that can be observed within an individual have clear direction (e.g., infection and onset of symptoms) but some may not (e.g., onset of infectiousness and onset of symptoms).
When we measure an epidemiological time delay between infected individuals (e.g., the serial interval), 
the primary and secondary events are defined in terms of the direction of transmission:
the primary event refers to the event that occurs in the infector. 
Again, some of these delays are always positive (the infector is always infected before the infectee) and some are not (it is possible for the infectee to develop symptoms before the infector \citep{he2020temporal}).

@@

Both forward and backward perspectives
must yield identical \emph{measurement} (e.g., the length of the incubation
period of a given individual is the same whether measured forward from
the time of infection or backward from the time of symptom onset).
Consequently, if we restrict attention to individuals with a given
incubation period, say $\tau$, and we count how many became infected at time $\psymp$
or how many first showed symptoms at time $\ssymp=\psymp+\tau$ then we must obtain
the same number. In general, no matter how delays are distributed, if
$\mathcal P$ and $\mathcal S$ represent the sizes of primary and
secondary cohorts then we must have
\begin{equation}
\psize(\psymp) f_\psymp(\tau) = \ssize(\ssymp) b_\ssymp(\tau) \,.
\label{eq:match}
\end{equation}
Substituting $\psymp=\ssymp-\tau$, it follows that
\begin{equation}
b_\ssymp(\tau)= \frac{\psize(\ssymp-\tau) f_{\ssymp-\tau} (\tau)}{\ssize(\ssymp)}\,.
\label{eq:backward}
\end{equation}
If we are thinking of realized incubation periods, then the left hand side of
this equation is the probability density that an individual
who became symptomatic at time $\ssymp$ had an incubation period of
length $\tau$. From the right hand side, we see that this probability
density depends on how many individuals became infected at time
$\ssymp-\tau$ [$\psize(\ssymp-\tau)$], how many individuals became
symptomatic at time $\ssymp$ [$\ssize(\ssymp)$], and the probability density
that someone who was infected at time $\ssymp-\tau$ became symptomatic at
time $\ssymp$ [$f_{\ssymp-\tau}(\tau)$]. In general, Eq.~\eqref{eq:backward}
shows that the the backward delay distribution at reference time $\ssymp$
[$b_\ssymp(\tau)$] depends not only on the secondary cohort size at time
$\ssymp$ [$\ssize(\ssymp)$], but also on the primary cohort size at the
earlier time $\ssymp-\tau$ [$\psize(\ssymp-\tau)$] and the forward delay
distribution at that earlier time [$f_{\ssymp-\tau}(\tau)$].

There are multiple mechanisms that drive the changes in forward and backward delay distributions over time.
Typically, within-individual forward delay distributions are not affected by epidemic dynamics.
Some realized distributions, like incubation period distributions, are expected to be equivalent to their intrinsic distributions and remain invariant over the course of an epidemic.
Other realized distributions, like the distribution of time from symptom onset to testing, may change over the course of an epidemic due to changes in public health policies.
Between-individual forward delay distributions, such as generation- or serial-interval distributions, depend on epidemic dynamics;
for example, forward generation intervals contract during the outbreak because infected individuals are less likely to infect others later due to susceptible depletion \citep{champredon2015intrinsic}.

@@

The serial interval is defined as the time between when an infector becomes symptomatic and when an infectee becomes symptomatic \citep{svensson2007note}.
Previous studies have often expressed serial intervals $\tau$ in the form (\fref{diagram}A):
\begin{equation}
\tau = (\gtime + x_1) - x_0
\end{equation}
where $x_0$ and $x_1$ represent incubation periods of an infector and an infectee, respectively, and $\gtime$ represents the generation interval.
These studies assumed that $x_0$ and $x_1$ follow the same distributions and concluded that the serial and generation intervals have the same mean \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation};
however, this conclusion is based on the implicit assumption that incubation times, $x_0$ and $x_1$, as well as generation interval, $\gtime$, are intrinsic to individuals (and not dependent on disease dynamics) --- something that is true of forward but not backward incubation times.
We refer to this definition as the intrinsic serial interval (\fref{diagram}A).

@@

The total number of serial intervals can now be obtained by integrating over all possible infection times for the infector and the infectee:
\begin{equation}
\total (\psymp,\ssymp) = \int_{-\infty}^{\psymp} \int_{\pinf}^{\ssymp} \Rc (\pinf) i(\pinf) h_{\pinf}(\psymp-\pinf, \sinf - \pinf) \idist_{\sinf}(\ssymp - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
Then, the forward serial-interval distribution $f_\psymp(\tau)$ is given by the density of intervals of length $\tau$ starting at \psymp, relative to the total number of serial intervals starting at \psymp: 
\begin{equation}
f_\psymp(\tau) = 
\frac{\total(\psymp, \psymp+\tau)}{\int \total(\psymp, \psymp+\tau') d\tau'}.
\label{eq:fserial}
\end{equation}
Likewise, the backward serial-interval distribution $b_\ssymp(\tau)$ is given by the density of intervals of length $\tau$ ending at \ssymp, relative to the total number of serial intervals ending at \ssymp: 
\begin{equation}
b_\ssymp(\tau) = 
\frac{\total(\ssymp-\tau, \ssymp)}{\int \total(\ssymp-\tau', \ssymp) d\tau'}.
\label{eq:bserial}
\end{equation}
This framework allows us to understand changes in the realized serial intervals for any deterministic epidemic models.

@@

We illustrate changes in the forward and backward serial intervals over the course of an epidemic by applying it to a specific example of an epidemic model. 
We model disease spread with a renewal-equation model \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, roberts2007model, champredon2018equivalence}.
Ignoring births and deaths, changes in the proportion of susceptible individuals $S(t)$ and incidence of infection $i(t)$ can be written as:
\begin{equation}
\begin{aligned}
\frac{\mathrm{d}S}{\mathrm{d}t} &= - i(t)\\
i(t) &= \Ro S(t) \int_0^\infty i(t-\tau) \gdist(\tau) \dtau,
\end{aligned}
\label{eq:renewal}
\end{equation}
where \Ro is the basic reproduction number, and $\gdist(\tau)$ is the intrinsic generation-interval distribution (i.e., the forward generation-interval distribution of a primary case in a population where the number of susceptibles remains constant).
Then, the forward generation-interval for a cohort of individuals that were infected at time $\psymp$ follows \citep{champredon2015intrinsic}:
\begin{equation}
\gdist_\psymp (\tau) \propto \gdist(\tau) S(\psymp + \tau),
\end{equation}
which allows us to separate the joint probability distribution $h_\psymp$ of the forward incubation period and the forward generation-interval distribution as a product of the proportion of susceptible individuals $S$ and the joint probability distribution $h$ of the forward incubation period and the intrinsic generation intervals:
\begin{equation}
h_\psymp (x, \tau) \propto h(x, \tau) S(\psymp + \tau).
\end{equation}
We further assume that the forward incubation period distributions does not vary across cohorts over the course of an epidemic, as they represent the life history of a disease, and use $k$ without subscript instead. 
Then, we have:
\begin{equation}
\begin{aligned}
\idist(x) &= \int_0^\infty h(x, \tau) \dtau\\
\gdist(\tau) &= \int_0^\infty h(x, \tau) \dx
\end{aligned}
\end{equation}
Finally, the case reproduction for this model is defined as follows:
\begin{equation}
\Rc(t) = \Ro \int_0^\infty \gdist(\tau) S(t+\tau) \dtau.
\end{equation}
The forward and backward serial-interval distributions are then calculated by substituting these quantities into \eref{fserial} and \eref{bserial}.
We use this framework to illustrate how the realized epidemiological time distributions vary over the course of an epidemic and depend on the perspective.

@@

During the initial phase of an epidemic, the proportion susceptible remains approximately constant ($S(t) \approx S(0)$) and incidence of infection grows exponentially: $i(t)=i_0\exp(rt)$.
During this period, the intrinsic generation-interval distribution provides the correct link between the exponential growth rate $r$ and the initial reproduction number $\RR=\Ro S(0)$ based on the Euler-Lotka equation \citep{wallinga2007generation};
in this case, we focus on the estimates of the basic reproduction number \Ro (the value of \RR in a fully susceptible population, $S(t) \approx 1$):
\begin{equation}
\frac{1}{\Ro} = \int_0^\infty \exp(-r\tau) g(\tau) \dtau.
\end{equation}
Analogous to the intrinsic generation-interval distribution, 
forward serial-interval distributions, by definition, describe the renewal process of infection based on symptom onset dates.
Therefore, we expect the forward serial-interval distribution during the exponential growth phase --- which we refer to as the \emph{initial} forward serial-interval distribution $f_0$ --- to estimate the same value of \Ro for a given $r$ as the intrinsic generation-interval distribution (note, however, that the forward serial interval is not necessarily positive):
\begin{equation}
\frac{1}{\Ro} = \int_{-\infty}^\infty \exp(-r\tau) f_{0}(\tau) \mathrm{d} \tau,
\end{equation}
where the initial forward serial-interval distribution is given by:
\begin{equation}
f_{0}(\tau) \propto \int_{-\infty}^{0} \int_{\pinf}^{\tau} \exp(r \pinf) h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}
In the Appendix, we provide a mathematical proof of this relationship.
We further compare this with the estimate of \Ro based on the intrinsic serial-interval distribution $q(\tau)$:
\begin{equation}
\frac{1}{\Rintrinsic} = \int_{-\infty}^\infty \exp(-r\tau) q(\tau) \mathrm{d} \tau,
\end{equation}
where the intrinsic serial-interval distribution $q(\tau)$ does not depend on disease dynamics:
\begin{equation}
q(\tau) \propto \int_{-\infty}^{0} \int_{\pinf}^{\tau} h(-\pinf, \sinf - \pinf) \idist(\tau - \sinf) \, \mathrm{d}\sinf\,\mathrm{d}\pinf.
\end{equation}

@@

\begin{table}[!th]
\begin{center}
\begin{tabular}{|l|l|r|}
\hline
Parameter & Values & Source\\
\hline
Mean intrinsic incubation period & 5.5 days & \cite{lauer2020incubation} \\
SD intrinsic incubation period & 2.4 days & \cite{lauer2020incubation} \\
Mean intrinsic generation interval & 5 days & \cite{ferretti2020quantifying} \\
SD intrinsic generation interval & 2 days & \cite{ferretti2020quantifying} \\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Parameter values used for simulations.}
The intrinsic generation-interval distribution is parameterized using a log-normal distribution with log mean $\mu_G=1.54$ and log standard deviation $\sigma_G=0.37$.
The forward incubation period distribution is parameterized using a log-normal distribution with log mean $\mu_I=1.62$ and log standard deviation $\sigma_I=0.42$.
The joint probability distribution is modeled using a multivariate log-normal distribution with correlations (on the log scale) $\rho=-0.5, 0, 0.5$.
}
\end{table}
